---
title: 'Advanced Methods: GLM and LMER'
output:
github_document:
toc: yes
toc_float: yes
---

```{r}
source('StatsResourcesConfig.R')
```

In the (previous notebook)[https://zenkavi.github.io/training/html/CorrelationTTestAnova.nb.html] we looked at linear models. In those models the residauals wwere assumed to be independent and normally distributed. 


## Generalized linear models

- Number of variables: >=2  
- Type of variables: 1 DV of any type (e.g. binary responses), >=1 IV both continuos and categorical  
- Output: (standardized) regression coefficients, t value, df, p value  
- Typical plot: scatterplot, boxplot, barplot  

```{r}
m1 = lm(Petal.Width ~ Petal.Length+Species, iris)

iris %>%
  mutate(pred_pw = predict(m1)) %>%
  ggplot(aes(Petal.Width, pred_pw))+
  geom_point()+
  xlab("Actual")+
  ylab("Predicted")
```

```{r}
m2 = lm(vs ~ mpg, mtcars)

mtcars %>%
  mutate(pred_vs = predict(m2)) %>%
  ggplot(aes(factor(vs), pred_vs))+
  geom_point()+
  xlab("Actual")+
  ylab("Predicted")+
  geom_hline(yintercept=0, col="red")+
  geom_hline(yintercept=1, col="red")
```

## Repeated measures

Within versus between subject factors
Fixed versus random effects

### ANOVA

### LMER