---
title: 'Conceptual Questions'
output:
github_document:
toc: yes
toc_float: yes
---

```{r}
source(StatsResourcesConfig.R)
```

## Conceptual questions

### What is a z-value

How many *population* standard deviations a raw value is away from the *population* mean.

```{r}
data = data.frame(x = rnorm(1000, mean=5, sd=1))
data = data %>%
  mutate(vals = dnorm(x, mean=5))

data %>%
  ggplot(aes(x, vals))+
  geom_line(group=1)
```

### What is a t-value

How many *sample* standard deviations a raw value is away from the *sample* mean.

```{r}
sample_data = sample_n(data, 20)

sample_data %>%
  ggplot()+
  geom_line(aes(x, vals), group=1, col='red')+
  geom_line(data=data, aes(x, vals), linetype='dashed')
```

```{r}
raw_value = sample_data$x[1]
cat("Sampled data point\n")
raw_value

cat("Mean of population\n")
mean(data$x)
cat("Standard deviation of population\n")
sd(data$x)

cat("z_score of sampled data point\n")
(raw_value - 5)/(1)

cat("Probability of sampling a value smaller than this number from this population\n")
pnorm((raw_value - 5)/(1))

cat("Mean of sample of 20\n")
mean(sample_data$x)
cat("Standard deviation of sample of 20\n")
sd(sample_data$x)

cat("t_score of sampled data point\n")
(raw_value - mean(sample_data$x)) / (sd(sample_data$x))

cat("Probability of sampling a value smaller than this number from this sample\n")
pt((raw_value - mean(sample_data$x)) / (sd(sample_data$x)), df = nrow(sample_data) -1)
```

### What is a p-value

Probability of observing your data given your null hypothesis. The smaller is it the more likely you are to 'reject' your null hypothesis. Note that this DOES NOT tell you   
- the probability that you alternative hypothesis is true or  
- the probability that your null hypothesis is false.  
To estimate probabilities of hypotheses depending on data you have observed you'd use Bayes' rule.